### 2017-04-05 paper/sexSelectionSpacing-ver1.tex
Added condensed discussion of why reversal of sex selection

The extension of birth spacing that result from abortions as described above is 
only one part of the interaction between son preference and birth spacing.

- the decision on sex selection and spacing
    - the mechanical extension
    - the possibility of changing use of sex selection

- what other factors might change birth spacing?
    - increased use of contraceptives
    - changes in labor force participation

@@

I show that there are cases where parents reconsider their decision to 
use sex selection between births, which underscores why 
using only observed births may provide a biased view of how much
sex selection takes place and what the final sex ratio will be.

As discussed above, shorter spacing after the birth of a girl is 
associated with worse health outcomes for both mother and children.%
\footnote{
\citet{wolpin84} and \citet{Newman1988} present dynamic models 
where both births and deaths are stochastic, but the likelihood of
a child dying is independent of the length of spacing to
the prior birth.
}
Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing is too short.

With the introduction of sex-selective abortions, ``too long'' spacing 
may, however, become a greater concern for parents, as 
repeated abortions can result in very long durations between births.%
\footnote{
Shorter spacing was probably less of a concern to begin with for
better educated mothers---who are also the most likely to use 
sex selection---since their children are substantially less likely 
to be negatively affected by short spacing than children of less 
educated mothers \citep{Whitworth2002}.
In general, mortality risk is negatively related to maternal
education \citep[for India, see, for example,][]{Tulasidhar1993},
although the underlying mechanism is subject
to debate \citep{Kovsted2002}.
}
If ``too long'' spacing leads parents to reverse their decision to use
prenatal sex determination within a spell this has important implications 
for the empirical method.
A reversal may happen, for example, if parents have concerns about possible 
infertility from repeated abortions.
Furthermore, with the increasing education and wealth in India, a growing 
portion of the population is likely to behave in a manner that makes the 
literature on shorter spacing in middle- and high-income countries more relevant.%
\footnote{
For the first birth, a potentially important factor in India is 
the pressure to show fecundity, which would tend to shorten
the duration to the first birth \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}
There may, for example, be economies of scale in childrearing---both
in terms of time cost and direct cost---that are easier to
take advantage of if births are closer \citep[p 947]{Newman1984}.
Similarly, if women's wages increase with age parents prefer
to have their children early and close together while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}

@@

To illustrate how the decision process works, consider
the decision in the final period on whether to use sex selection or not
(assuming that both are superior to no birth).
Parents choose a pregnancy without prenatal sex determination
if $U_T^B > U_T^S$, which leads to
\begin{multline} \label{eq:last_simple}
(1-\pi)(1-\alpha) [ u^g(g_{T-1}+1) - u^g(g_{T-1}) ]
> \\
\pi u(I - p_k(b_{T-1} + g_{T-1} + 1) - p_s) 
+ (1-\pi) u(I - p_k(b_{T-1} + g_{T-1}) - p_s) \\
- u(I - p_k(b_{T-1} + g_{T-1}+1)).
\end{multline}
Hence, parents choose a pregnancy without prenatal scan over
a pregnancy with a prenatal scan in the final period if
the additional utility of having an extra girl (the left-hand-side),
weighted by the probability of having a girl and the weight parents
assign to utility from having girls, is larger than the expected impact on
utility from consumption (the right-hand side), where
the first two terms are the expected utility from consumption if 
using prenatal sex determination and the last term is utility from
consumption if parents simply have a birth.%
\footnote{
Notice that the marginal utility of having a boy does not enter
into the decision on whether to use prenatal sex determination or not
(except through the value of $\alpha$).
Only the marginal utility of having another girl
and the total number of children, which affects the trade-off between
consumption and use of sex selection, matters.
This is an artifact of the separability between consumption, number of
boys, and number of girls in the utility function.
}

If a son is conceived the money for a prenatal scan are ``wasted,''
which means that the utility from consumption if they use prenatal scan
and conceive a boy is alway going to be lower than utility from consumption 
if the parents had a birth without a prenatal scan.
Hence, assuming that the marginal utility of girls is always positive,
but declining in the number of girls they have, parents will pick
a birth with scan over a birth without scan only if the difference between
utility from consumption when using prenatal scan and conceiving a girl
is sufficiently higher than the utility of consumption if the parents
simply had a child without a scan.
It follows that the higher the cost of the prenatal scan is relatively
to the cost of maintaining a child the less likely it is that parents will 
use sex selection.
Furthermore, if the cost of a prenatal scan, 
$p_s$, is larger than the cost of maintaining a child in one period, $p_k$, 
parents will never select a prenantal scan in the final period.
This result holds, however, only because this is the last period. 
The earlier the decision on sex selection is made in the time horizon
the lower the cost of prenatal sex determination is going to be 
relative to the cost of maintaining the child until the end of the time
horizon.

>>>>

In the absence of prenatal sex determination, parents' only choice if they 
wanted a son was to try for another pregnancy as soon as possible after
the birth of a girl. 
This led to shorter spacing after the birth of a girl and what
is known as differential stopping behavior, where parents were 
more likely to cease child bearing after the birth of a son than
after a daughter [TK add references here]. 
As discussed above, the shorter spacing after the birth of a girl is 
associated with worse health outcomes for both mother and children.
Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing was too short.

With the introduction of sex-selective abortions, parents can abort
fetuses of unwanted sex.
With an approximately 12 months additional duration per abortion,
this strategy can result in very long durations between births, and
``too long'' spacing may become a greater concern for parents than 
``too short'' spacing.%
\footnote{
Shorter spacing was likely less of a concern to begin with for
better educated mothers---who are also the most likely to use 
sex selection---since their children are substantially less likely 
to be negatively affected by short spacing than children of less 
educated mothers \citep{Whitworth2002}.
}
If the duration from last birth becomes long enough parents may 
reverse their decision to use prenatal sex determination
and carry the next pregnancy to term whether it is a male or female fetus.

At least four different factors may make a reversal of the decision 
to use sex selection between one birth and the next more likely, 
everything else equal.
First, the more concerned parents are about possible infertility 
from repeated abortions or the closer they are to the end of their 
reproductive period, they more likely they are to decide that any 
child is better than not having an additional child at all.
Second, the higher the parents' discount factor is---that is, the
more the value present outcomes relative to future outcomes---the
more likely they are to reverse their decision to sex selection.
Third, economies of scale in childrearing---both in terms of time cost 
and direct cost---are easier to take advantage of if births 
are closer \citep[p 947]{Newman1984}.
Hence, the stronger these economies of scale are, the more likely
parents are to reverse their decision to use sex selection.
Finally, if women's wages increase with age, parents prefer
to have their children early and close together, while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}
The more pronounced the increase in wage is with age, the more 
attractive it will be to reverse the decision to use sex selection.
Although, the latter two are most associated with the literature
on shorter spacing in middle- and  high-income countries, 
the increasing education and wealth in India means that a growing 
portion of the population is likely to behave in a similar manner.%
\footnote{
The possible exception is for the first birth, where the pressure to show 
fecundity---which shortens the duration to the first birth---is potentially 
an important factor in India \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}

If ``too long'' spacing leads parents to reverse their decision to use
prenatal sex determination within a spell this has important implications 
for the empirical method.

@@

@@
